---
title: "Untitled"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
installed_packages <- installed.packages()[,1]
isDplyrInstalled <- "dplyr" %in% installed_packages
isGgplotInstalled <- "ggplot2" %in% installed_packages
isReshape2Installed <- "reshape2" %in% install.packages

if(!isDplyrInstalled) {
  install.packages("dplyr")
}

if(!isGgplotInstalled) {
  install.packages("ggplot2")
}

if(!isReshape2Installed) {
  install.packages("reshape2")
}

library(dplyr)
library(ggplot2)
library(reshape2)

#Import uber data
uber_data <- read.csv("Uber Request Data.csv")

View(uber_data)

str(uber_data)

summary(uber_data)

#Make a copy of uber data
master_uberdata <-  uber_data

#Conversion of time format $d/$m/$Y %H:%M 
master_uberdata$ReqtimeFormatted <- as.POSIXct(master_uberdata$Request.timestamp, format="%d/%m/%Y %H:%M")

#Conversion of time format $d-$m-$Y %H:%M:%S
master_uberdata$Request.timestamp <- as.POSIXct(master_uberdata$Request.timestamp, format="%d-%m-%Y %H:%M:%S")

#Verify if there are any missing values. If the below expression is true then there are none of the req time is missed.
length(which(is.na(master_uberdata$Request.timestamp))) == length(which(!is.na(master_uberdata$ReqtimeFormatted)))

#Conversion of time format $d/$m/$Y %H:%M
master_uberdata$DroptimeFormatted <- as.POSIXct(master_uberdata$Drop.timestamp, format="%d/%m/%Y %H:%M")

#Conversion of time format $d-$m-$Y %H:%M:%S
master_uberdata$Drop.timestamp <- as.POSIXct(master_uberdata$Drop.timestamp, format="%d-%m-%Y %H:%M:%S")

#Verify if there are any missing values. If the below expression is true then there are none of the Drop time is missed.
length(which(is.na(master_uberdata$Drop.timestamp) & master_uberdata$Status == "Trip Completed")) == length(which(!is.na(master_uberdata$DroptimeFormatted) & master_uberdata$Status == "Trip Completed"))

#Merge fromated req time coulmns and drop time columns 
master_uberdata <- mutate(master_uberdata, Requesttime = as.POSIXct(ifelse(is.na(Request.timestamp),ReqtimeFormatted,Request.timestamp), origin = "1970-01-01"))

master_uberdata <- mutate(master_uberdata, Droptime = as.POSIXct(ifelse(is.na(Drop.timestamp),DroptimeFormatted,Drop.timestamp), origin = "1970-01-01"))


#Remove unwnated rows
master_uberdata <- master_uberdata[c(1,2,3,4,9,10)]

#Genereate Weekdays derived variable
master_uberdata$Weekdays <- factor(weekdays(master_uberdata$Requesttime), levels = c("Monday","Tuesday","Wednesday","Thursday","Friday"))

#Genereate Travel time derived variable
master_uberdata$Traveltime <- (master_uberdata$Droptime - master_uberdata$Requesttime)
master_uberdata$Traveltime <-round(as.numeric(master_uberdata$Traveltime), digits=0)

#Genereate Hour derived variable
master_uberdata$Hour <- as.numeric(format(master_uberdata$Requesttime, "%H"))

#Calculate request cancelled, success and no cars available
master_uberdata <- group_by(master_uberdata,Status)
Status_summary <- summarise(master_uberdata, Request_Count = length(Request.id))
#Group by weekday and status
master_uberdata <- group_by(master_uberdata, Weekdays, Status)
weekday_Status_summary <- summarise(master_uberdata, BookingStatus = length(Weekdays))
```
```{r pressure, echo=FALSE}
#plot the Bookings w.r.t hours
ggplot(master_uberdata, aes(x=Hour, fill=Status)) + geom_bar(position = "dodge") + 
  labs(title="Hour vs Request Count(All the Weekdays together)", y="Request Count", X= "Hours")

#Plot the Bookings w.r.t each day along with the Status fo the booking
ggplot(master_uberdata, aes(x=Weekdays, fill=Status)) + geom_bar(position = "dodge") + 
  labs(title = "Weekdays vs Request", y="No of Request")
```

```{r cars}
#Generate a column with Timslot containin evening, morning, Noon, Night, Midnight, Early morning
CalculatetimeSlot <- function(hour){
                      if( 3 < hour & hour <=  7)
                      {
                        return("Early Morning")
                      }
                      if( 7 < hour & hour <= 11 )  
                      {
                        return("Morning")
                      }  
                      if(11 < hour & hour <= 15 )  
                      {
                        return("Noon")
                      } 
                      if( 15 < hour & hour <= 19 )  
                      {
                        return("Evening")
                      }
                      if( 19 < hour & hour <= 23 )  
                      {
                          return("Night")
                      }
                      if(hour > 23  | hour <= 3)  
                      {
                          return("Midnight")
                      }
                      return("Other")
                  }
master_uberdata$TimeSlot <- c(NA)
for(i in 1:nrow(master_uberdata))
{    
  master_uberdata$TimeSlot[i] <- CalculatetimeSlot(master_uberdata$Hour[i])
}
master_uberdata$TimeSlot <- factor(master_uberdata$TimeSlot, levels=c("Early Morning","Morning","Noon","Evening","Night","Midnight"))

#including timeslot in gorup by
master_uberdata <- group_by(master_uberdata, Weekdays, TimeSlot, Status)
time_summarry <- summarise(master_uberdata, No_of_request = length(Weekdays))
```
```{r pressure, echo=FALSE}
#Plot graph with Status, Timslot, Weekdays, No of Request

ggplot(time_summarry, aes(x=TimeSlot, y=No_of_request, fill=Status)) + geom_col(position="dodge") + 
     facet_grid(.~Weekdays) + labs(title = "Weekdays, Timeslot vs Request", y="No of Request") +
     theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggplot(time_summarry, aes(x=TimeSlot, y=No_of_request, fill=Status)) + geom_col(position="dodge") + 
  labs(title = "Timeslot vs Request", y="No of Request") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#From the graph we could find that the no cars available and cancelled is high during evening and early morning respectively
```

```{r cars}
#Filtering the uber data with no car available and evening timezone
evening_nocars <- filter(master_uberdata, TimeSlot == "Evening", Status == "No Cars Available")
evening_nocars <- group_by(evening_nocars, Pickup.point, Weekdays)
nocarsreq_weekday <- summarise(evening_nocars, No_Of_Request = length(Weekdays))
nocarsreq_weekday$Weekdays <- factor(nocarsreq_weekday$Weekdays, levels = c("Monday","Tuesday","Wednesday","Thursday","Friday"))
```
```{r pressure, echo=FALSE}
#Plot Cars not available req vs pickup point by weekday
ggplot(nocarsreq_weekday, aes(x=Pickup.point, y=No_Of_Request, fill=Weekdays)) + geom_col(position="dodge") + 
  geom_text(aes(label=No_Of_Request), vjust = 1, position = position_dodge(width=1)) +
  labs(title = "Pickup point vs No of Request (Cars not available during Evening by Day)", y="No of Request")

#Plot Cars not available vs pickup point by Hour
evening_nocars <- group_by(evening_nocars, Pickup.point, Hour)
nocarsreq_Hour <- summarise(evening_nocars, No_Of_Request = length(Hour))
ggplot(nocarsreq_Hour, aes(x=Pickup.point, y=No_Of_Request, fill=as.factor(Hour))) + geom_col(position="dodge") + 
  geom_text(aes(label=No_Of_Request), vjust = 1, position = position_dodge(width=1)) +
  labs(title = "Pickup point vs No of Request (Cars not available during evening by Hour)", y="No of Request")
```
#Filtering the uber data with cancelled and early morning
earlymrng_cancelled <- filter(master_uberdata, TimeSlot == "Early Morning", Status == "Cancelled")
earlymrng_cancelled <- group_by(earlymrng_cancelled, Pickup.point, Weekdays)
cancelreq_weekday <- summarise(earlymrng_cancelled, No_Of_Request = length(Weekdays))

#Plot cancel req vs pickup point by weekday
ggplot(cancelreq_weekday, aes(x=Pickup.point, y=No_Of_Request, fill=Weekdays)) + geom_col(position="dodge") + 
  geom_text(aes(label=No_Of_Request), vjust = 1, position = position_dodge(width=1)) +
  labs(title = "Pickup point vs No of Request (Cancelled During Early morning by Day)", y="No of Request")

#Plot cancel req vs pickup point by Hour
earlymrng_cancelled <- group_by(earlymrng_cancelled, Pickup.point, Hour)
cancelreq_Hour <- summarise(earlymrng_cancelled, No_Of_Request = length(Hour))
ggplot(cancelreq_Hour, aes(x=Pickup.point, y=No_Of_Request, fill=as.factor(Hour))) + geom_col(position="dodge") + 
  geom_text(aes(label=No_Of_Request), vjust = 1, position = position_dodge(width=1)) +
  labs(title = "Pickup point vs No of Request (Cancelled During Early morning by Hour)", y="No of Request")

#Calculating supply demand gap  Count of cancelled and no cars available tips - count of tips completed
airport_req <- filter(master_uberdata, Pickup.point == "Airport")
airport_req <- group_by(airport_req, TimeSlot)
airport_supply_demand_gap <- summarise(airport_req, Demand = length(which(!is.na(Request.id))), Supply = length(which(Status=="Trip Completed")), 
                               Supply_Demand_Gap = (length(which(!is.na(Status))) - length(which(Status=="Trip Completed"))))

air_gap_melt <- melt(airport_supply_demand_gap, id.vars='TimeSlot')

#Plot graph for supply  and demand for pickup points air
ggplot(air_gap_melt, aes(x=TimeSlot, y=value, fill=variable)) + geom_bar(stat='identity', position='dodge') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(title = "Supply Demand Gap for pickup points with Airport", y="Request")


#Calculating supply demand gap  Count of cancelled and no cars available tips - count of tips completed
city_req <- filter(master_uberdata, Pickup.point == "City")
city_req <- group_by(city_req, TimeSlot)
city_supply_demand_gap <- summarise(city_req, Demand = length(which(!is.na(Request.id))), Supply = length(which(Status=="Trip Completed")), 
                                       Supply_Demand_Gap = (length(which(!is.na(Status))) - length(which(Status=="Trip Completed"))))

city_gap_melt <- melt(city_supply_demand_gap, id.vars='TimeSlot')

#Plot graph for supply  and demand for pickup points air
ggplot(city_gap_melt, aes(x=TimeSlot, y=value, fill=variable)) + geom_bar(stat='identity', position='dodge') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(title = "Supply Demand Gap for pickup points with City", y="Request")

#Analyse why there are cancel and cars not available

#Filter success trips details
success_trips <- filter(master_uberdata, !is.na(Traveltime))

#Plot graph for Traveltime
success_trips_airport <- filter(master_uberdata, Pickup.point == "Airport")
ggplot(success_trips_airport, aes(x=TimeSlot, y=Traveltime)) + geom_boxplot(notch = TRUE, aes(colour = TimeSlot)) + 
  labs(title="Traveltime") + geom_jitter(aes(colour=TimeSlot, shape = TimeSlot))

#calculate median and quartile for travel time duirng evening slot
success_trips_airport_evening <- filter(success_trips_airport, TimeSlot == "Evening")
summary(success_trips_airport_evening)

#Filter the round trips
success_trips <- arrange(success_trips,Driver.id,Requesttime, Droptime)
roundtrip_index <- c()
for(i in 1:nrow(success_trips))
{
   if(i == nrow(success_trips))
   {
     next
   }   
   if((success_trips$Pickup.point[i] != success_trips$Pickup.point[i+1]) & (success_trips$Driver.id[i] == success_trips$Driver.id[i+1]))
   {
     roundtrip_index[length(roundtrip_index) + 1] <- i
   }
}

round_trips <- success_trips[roundtrip_index,]

#Calculate Idle time
round_trips$idleTime <- c(NA)
for(i in 1:nrow(round_trips))
{
  if(i == nrow(round_trips))
  {
    next
  } 
  if(round_trips$Driver.id[i] == round_trips$Driver.id[i+1])
    round_trips$idleTime[i] <- round_trips$Requesttime[i+1] - round_trips$Droptime[i] 
}


#Considering only the idle time in airport
round_trips_city <-  filter(round_trips, Pickup.point == "City")
round_trips_city$idleTime <- round(round_trips$idleTime,0)

#Plotting the idle time
ggplot(round_trips_city,aes(x=TimeSlot, y=idleTime)) + geom_boxplot(notch = TRUE, aes(colour = TimeSlot)) + 
  labs(title="Idle Time") + geom_jitter(aes(colour=TimeSlot, shape = TimeSlot))

#calculate median and quartile for idle time duirng early morning slot
round_trips_city_morning <- filter(round_trips_city, TimeSlot == "Early Morning")
summary(round_trips_city_morning)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
